A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-04-19 graph_syntax.tex
fixing a few last things

\section{Syntax Definition}
  \label{s:syntax definition}
  In this section, we present the necessary graph theoretic fundamentals to 
  construct the graphs discussed in Sec.~\ref{s:requirements}. 
  The notation used in this work is adapted from Diestel \cite{Diestel2010}. 
  A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
  which means that the elements of $E$ are 2--element subsets of $V$. The set $V$ 
  contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
  For a \emph{directed graph*}((DJP: why the asterisk?)) (or \emph{digraph}) we construct $E$ as a set of ordered pairs instead 
  of a set of sets. Each ordered pair represents an edge starting at the node 
  indicated by the first entry and directed to the node indicated by the second 
  entry. Edge $e$ = $(x,y)$ may be referred to simply as $xy$. For node $v \in V$ 
  the edges directed out from $v$ are denoted by $E(v)$ and the edges directed into $v$ are given 
  by $E^{-1}(v)$. 
  If $E$ is not a one--to--one mapping, $E(v)$ may be the empty set, a single node, or a set of nodes.
  \begin{figure}[htb!]
    \begin{center}
    \includegraphics[width=1.5in]{images/example_directed_graph}
    \end{center}
    \vspace{-20pt}
  \caption{Example directed graph.}
  \label{f:example directed graph}
  \end{figure}
  As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
  \begin{IEEEeqnarray*}{rCl}
  V & = & \{1,2,3,4,5\}, \\
  E & = & \big\{(2,1),(3,2),(1,3),(3,4)\big\}.
  \end{IEEEeqnarray*}

@@

\subsection{indegree and outdegree}
  \label{s:indegree-outdegree}
  The \emph{indegree} of a node is the number of connection edges directed in and 
  is denoted as $\txt{deg}^-(v)$, and the \emph{outdegree} 
  is the number of connection edges directed out and it is denoted as $\txt{deg}^+(v)$.
  The degree of a given node is a function only of the connection edges 
  attached to it. The number of driven edges is not relevant because any number 
  of drivers could be involved in different parts of an solution process. For 
  example, in a sequential optimization in which a global optimizer is applied first ((BG: first and second?))
  and a gradient based optimizer is applied second, design variables would have incomming 
  driven edges from both optimizers but this would not affect the indegree of those
  variable nodes.

>>>>

\section{Syntax Definition}
  \label{s:syntax definition}
  In this section, we present the necessary graph theoretic fundamentals to 
  construct the graphs discussed in Sec.~\ref{s:requirements}. 
  The notation used in this work is adapted from Diestel \cite{Diestel2010}. 
  A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
  which means that the elements of $E$ are 2--element subsets of $V$. The set $V$ 
  contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
  For a \emph{directed graph} (or \emph{digraph}) we construct $E$ as a set of ordered pairs instead 
  of a set of sets. Each ordered pair represents an edge starting at the node 
  indicated by the first entry and directed to the node indicated by the second 
  entry. Edge $e$ = $(x,y)$ may be referred to simply as $xy$. For node $v \in V$ 
  the edges directed out from $v$ are denoted by $E(v)$ and the edges directed into $v$ are given 
  by $E^{-1}(v)$. 
  If $E$ is not a one--to--one mapping, $E(v)$ may be the empty set, a single node, or a set of nodes.
  \begin{figure}[htb!]
    \begin{center}
    \includegraphics[width=1.5in]{images/example_directed_graph}
    \end{center}
    \vspace{-20pt}
  \caption{Example directed graph.}
  \label{f:example directed graph}
  \end{figure}
  As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
  \begin{IEEEeqnarray*}{rCl}
  V & = & \{1,2,3,4,5\}, \\
  E & = & \big\{(2,1),(3,2),(1,3),(3,4)\big\}.
  \end{IEEEeqnarray*}

@@

\subsection{indegree and outdegree}
  \label{s:indegree-outdegree}
  The \emph{indegree} of a node is the number of connection edges directed in and 
  is denoted as $\txt{deg}^-(v)$, and the \emph{outdegree} 
  is the number of connection edges directed out and it is denoted as $\txt{deg}^+(v)$.
  The degree of a given node is a function only of the connection edges 
  attached to it. The number of driven edges is not relevant because any number 
  of drivers could be involved in different parts of an solution process. For 
  example in a sequential optimization in which a global optimizer is used first 
  followed by a gadient based optimizer second to refine global optimizer 
  result, design variables would have incomming driven edges from both optimizers. 
  These driven edges would not affect the indegree of those variable nodes.

### 2013-04-19 intro.tex
fixing a few last things

    The time and expense for setting up 
    analysis models is growing in conjunction with the increasing scope of MDO problems. 
    Multidisciplinary Design Analysis and Optimization (MDAO)
    frameworks such as OpenMDAO\cite{Gray2012} and ModelCenter ((BG: TM?)) have enabled a new level of analysis tool integration 
    and have paved the way for models with more analyses and increasing numbers of interdisciplinary couplings. That 
    new capability has created a new challenge. Even models with tens of analyses could include hundreds or thousands
    of variables that are interdependent and must be linked in the framework. 
    A common situation is that different analyses provide differing values for the 
    same physical quantity, and these conflicts need to be resolved. These differing values 
    may not even be in the same format. These occurrences are particularly acute when analysis tools have differing fidelities. For
    example, an abstracted aerodynamic analysis such as an empirical drag buildup model may return only
    integrated drag, whereas a CFD tool may return pressure and shear stress distributions across the entire
    surface grid. If an analysis downstream of the aerodynamics tool needs only integrated drag as an input,
    then the designer has a free choice of which of the two possible aerodynamics analysis tools to select to
    provide the drag estimate (presuming that drag can be computed from surface distributions by a simple
    integration algorithm). On the other hand, if a downstream analysis needs a pressure distribution in
    order to compute pitching moment, for example, then any feasible MDO problem formulation must
    include CFD or similar analysis in the data flow, regardless of whether the empirical drag buildup model
    is also included.

@@

    A DSM by itself is insufficient to describe complete problem formulations because it captures only information about data dependency between analyses; objective and constraint information is missing from the DSM description of the problem. 
    An alternative matrix--based syntax, called a Functional Dependency Table (FDT), was proposed by Wagner and Papalambros. 
    FDT represents the relationship between functions, including objectives and constraints, and specific variables that affect 
    them\cite{Wagner1993}. Similar to DSM, FDT also describes an adjacency matrix of a graph. Unlike the DSM graph, 
    however, the FDT graph is undirected and nodes can represent analysis tools, objectives, 
    or constraints. Edges between nodes represent a dependence on the same 
    variable. Michelena and Papalambros made use of the FDT to solve a graph partitioning problem that yielded 
    more efficient optimization problem decompositions\cite{Michelena1997}. While FDT succeeds at capturing the 
    information about objectives and constraints, its lack of directed edges implies that it cannot capture the coupled data dependency that ((DJP: \emph{a}?)) DSM captures. For example, 
    the objective, $f$, from the FPF in Eq. \ref{eqn:simple_fpf} is dependent on 
    output $m$ of analysis $A$. The dependence of $m$ on $A$ is now captured 
    in FDT in Fig. \ref{fig:FDT_simple} alone. This missing information implies 
    that, while FDT is very useful for partitioning problems, it is also not 
    sufficient to contain a complete problem formulation.

>>>>

    The time and expense for setting up 
    analysis models is growing in conjunction with the increasing scope of MDO problems. 
    Multidisciplinary Design Analysis and Optimization (MDAO)
    frameworks such as OpenMDAO\cite{Gray2012} and \textregistered{ModelCenter} have enabled a new level of analysis tool integration 
    and have paved the way for models with more analyses and increasing numbers of interdisciplinary couplings. That 
    new capability has created a new challenge. Even models with tens of analyses could include hundreds or thousands
    of variables that are interdependent and must be linked in the framework. 
    A common situation is that different analyses provide differing values for the 
    same physical quantity, and these conflicts need to be resolved. These differing values 
    may not even be in the same format. These occurrences are particularly acute when analysis tools have differing fidelities. For
    example, an abstracted aerodynamic analysis such as an empirical drag buildup model may return only
    integrated drag, whereas a CFD tool may return pressure and shear stress distributions across the entire
    surface grid. If an analysis downstream of the aerodynamics tool needs only integrated drag as an input,
    then the designer has a free choice of which of the two possible aerodynamics analysis tools to select to
    provide the drag estimate (presuming that drag can be computed from surface distributions by a simple
    integration algorithm). On the other hand, if a downstream analysis needs a pressure distribution in
    order to compute pitching moment, for example, then any feasible MDO problem formulation must
    include CFD or similar analysis in the data flow, regardless of whether the empirical drag buildup model
    is also included.

@@

    A DSM by itself is insufficient to describe complete problem formulations because it captures only information about data dependency between analyses; objective and constraint information is missing from the DSM description of the problem. 
    An alternative matrix--based syntax, called a Functional Dependency Table (FDT), was proposed by Wagner and Papalambros. 
    FDT represents the relationship between functions, including objectives and constraints, and specific variables that affect 
    them\cite{Wagner1993}. Similar to DSM, FDT also describes an adjacency matrix of a graph. Unlike the DSM graph, 
    however, the FDT graph is undirected and nodes can represent analysis tools, objectives, 
    or constraints. Edges between nodes represent a dependence on the same 
    variable. Michelena and Papalambros made use of the FDT to solve a graph partitioning problem that yielded 
    more efficient optimization problem decompositions\cite{Michelena1997}. While FDT succeeds at capturing the 
    information about objectives and constraints, its lack of directed edges 
    implies that it cannot capture the coupled data dependency that a DSM captures. For example, 
    the objective, $f$, from the FPF in Eq. \ref{eqn:simple_fpf} is dependent on 
    output $m$ of analysis $A$. The dependence of $m$ on $A$ is now captured 
    in FDT in Fig. \ref{fig:FDT_simple} alone. This missing information implies 
    that, while FDT is very useful for partitioning problems, it is also not 
    sufficient to contain a complete problem formulation.
